{Growing} a Static Type System Inside a Python

### 2016-06-17 atlang-modularity15/atlang-modularity15.tex
progress on GPCE paper

%\documentclass[12pt]{article}
\documentclass[9pt,preprint]{sigplanconf}
\newcommand{\sigmat}[1]{\sigma_{\text{#1}}}

\newcommand{\F}[1]{{\sf #1}~}
\newcommand{\FF}[1]{{\sf #1}}
\newcommand{\Q}{\FF{Arg}}
\usepackage{sfmath}
\usepackage{times}
%\setlength{\parskip}{0pt}
%\setlength{\parsep}{0pt}
\setlength{\headsep}{10pt}
%\setlength{\topskip}{0pt}
%\setlength{\topmargin}{0pt}
%\setlength{\topsep}{0pt}
%\setlength{\partopsep}{0pt}

@@

\newcommand{\lstinlinep}[1]{\lstinline[language=Python,basicstyle=\ttfamily\small,deletendkeywords={tuple,buffer,map}]{#1}}

@@

\lefthyphenmin=8
\sloppy
%\makeatletter
%\renewcommand\section{\@startsection{section}{1}{\z@}%
%                       {-18\p@ \@plus -4\p@ \@minus -4\p@}%
%                       {8\p@ \@plus 4\p@ \@minus 4\p@}%
%                       {\normalfont\large\bfseries\boldmath
%                        \rightskip=\z@ \@plus 8em\pretolerance=10000 }}
%\renewcommand\subsection{\@startsection{subsection}{2}{\z@}%
%                       {-18\p@ \@plus -4\p@ \@minus -4\p@}%
%                       {8\p@ \@plus 4\p@ \@minus 4\p@}%
%                       {\normalfont\normalsize\bfseries\boldmath
%                        \rightskip=\z@ \@plus 8em\pretolerance=10000 }}
%\renewcommand\subsubsection{\@startsection{subsubsection}{3}{\z@}%
%                       {-5\p@ \@plus -4\p@ \@minus -4\p@}%
%                       {-0.5em \@plus -0.22em \@minus -0.1em}%
%                       {\normalfont\normalsize\bfseries\boldmath}}
%\renewcommand\paragraph{\@startsection{paragraph}{4}{\z@}%
%                       {-4\p@ \@plus -4\p@ \@minus -4\p@}%
%                       {-0.5em \@plus -0.22em \@minus -0.1em}%
%                       {\normalfont\normalsize\itshape}}
%\makeatother

@@

\conferenceinfo{-}{-} 
\copyrightyear{-} 
\copyrightdata{[to be supplied]}

\title{Extensible Gradual Typing Inside a Python}
%\subtitle{Unambiguously Growing a Static Type System Inside a Python}
%\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
%\preprintfooter{Distribute within CMU only.}   % 'preprint' option specified.

%\authorinfo{~}{~}{~}
\authorinfo{Cyrus Omar \and Jonathan Aldrich}
          {          Carnegie Mellon University}
          {\{comar, aldrich\}@cs.cmu.edu}

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
%gain access to new abstractions, even when they agree that these abstractions
%could be valuable. This suggests a need for languages that satisfy two key
%growth criteria: compatibility with existing programming ecosystems; and
%internal extensibility, so that new abstractions do not require new
%languages.
%
%We introduce @\texttt{lang}, a language compatible with the popular Python ecosystem.
%While Python, like other similar languages, is satisfactory for simple
%scripting, @\texttt{lang} is designed for more complex situations where static
%typechecking is beneficial. Unlike most statically-typed languages, @\texttt{lang}'s type
%system can be extended from within, by associating compile-time functions with
%type definitions. The compiler selectively invokes these according to a
%type-directed protocol, avoiding interference problems. Despite these design
%constraints, active types in @\texttt{lang} can express the static and dynamic semantics
%of a range of functional, object-oriented, parallel and domain-specific
%abstractions, all within libraries.
Programmers prefer using abstractions available as libraries for the language they already use over those requiring a switch to a new language. But library providers cannot typically extend the static type systems of these languages directly, so powerful statically typed abstractions  remain obscure. This suggests the need for a mechanism that allows abstraction providers to extend the type system of a widely adopted language from within, in a way that guarantees that any combination of extensions can be used together without worrying about conflicts.

Our aim in this paper is to 1) introduce such a mechanism; and 2) show how it can be implemented as a library for an existing widely used language, here Python. We describe two artifacts: @\texttt{lang}, an  extensible statically typed language implemented as a library within Python, and @$\lambda$, a minimal core calculus. Both @\texttt{lang} and @$\lambda$ have a fixed syntax. They determine the semantics of each term by delegating control, according to a simple but powerful bidirectionally typed elaboration semantics, to methods associated with a relevant \emph{active type constructor}, defined in a library. This process is stable even when additional extensions are imported, so we avoid syntactic and semantic ambiguity by construction. We validate our approach with several examples of interesting type system fragments that are built into other languages but can be implemented as libraries using our approach, including functional constructs, a type system for secure string sanitation and a typed foreign function interface to OpenCL.% By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction. %In @\texttt{lang}, types and functions are defined using Python's syntax, distributed using Python's package system and can call into Python libraries, so @\texttt{lang} is uniquely backwards compatible. We show that active type constructors as implemented in @\texttt{lang} permit practical library implementations of a variety of type system fragments that would normally need to be built into a language. %We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into @\texttt{lang} and then through to OpenCL. % can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library. %We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular. %@\texttt{lang} can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces @\texttt{lang}, a language that aims to satisfy this more comprehensive set of design criteria. To do so, @\texttt{lang} introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of @\texttt{lang} is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}\label{intro}
Asking programmers to import a library is easier than asking them to adopt a new programming language. Indeed, recent empirical studies underscore the difficulties of driving new languages into adoption, finding that extrinsic factors like compatibility with large existing code bases, library availability, familiarity and tool support are at least as important as intrinsic features of the language \cite{Meyerovich:2013:EAP:2509136.2509515,chen05}.

Unfortunately, researchers and domain experts designing new abstractions sometimes find it difficult to soundly, completely and naturally implement them as libraries, particularly those that rely on specialized support from a static type system. In these situations, abstraction providers generally develop a new language dialect, increasingly with the aid of tools like language workbenches \cite{erdweg2013state}. Unfortunately, this \emph{language-oriented approach} \cite{journals/stp/Ward94} is problematic at language boundaries: working with components written in several distinct languages often requires programming against their low-level implementation details, rather than their semantics, via a foreign-function interface (FFI) to a common implementation language (e.g. C or the JVM). This is brittle, unnatural and unsafe and can cause performance issues due to necessary checks and conversions at language boundaries. As more languages are used together, more FFIs need to be constructed and traversed.%To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec. \ref{related-work}).

These issues prevent even those dialects that introduce only a few new constructs from being adopted. A study comparing a Java dialect, Habanero-Java (HJ), with a comparable library, \lstinlinep{java.util.concurrent}, for example, found that the language-based abstractions in HJ were easier to use and provided useful static guarantees \cite{cave2010comparing}. Nevertheless, it concluded that the library-based abstractions remained more practical outside the classroom because using HJ, as a distinct dialect of Java with its own type system, would be problematic in settings where some developers had not adopted it, but needed to interface with code that had. It would also be difficult to use it in combination with other abstractions also implemented as dialects of Java. Moreover, its tool support is more limited. 
This appears to be a widespread problem: programmers and development teams often cannot use  the typed abstractions they prefer because they are only available bundled with specialized languages they cannot  adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}. %These languages become, in essence, merely vehicles for communicating ideas, rather than useful tools.

@@

Extensible languages have promised to reduce the problems associated with distinct dialects by giving abstraction providers more direct and granular control over a language's syntax and semantics from within, or alongside, libraries. %In   an internally extensible language, programmers  can \emph{granularly} import the primitive constructs best suited to each component of their application or library. Providers can thus more easily develop, deploy and evaluate new abstractions in the context of existing codebases, narrowing one of the gaps between research and practice \cite{basili2008understanding}. 
%Indeed, abstractions could compete on their individual merit, rather than  by the serious adoption barriers mentioned above.
Unfortunately, the mechanisms available today have several problems. First, they are themselves often available only within a dialect of an existing language and so face the classic ``chicken-and-egg'' problem: a language like SugarJ \cite{erdweg2011sugarj} or Wyvern \cite{TSLs} (which permit only syntax extensions over a fixed semantics) must overcome the same extrinsic barriers as a language like HJ. Second, giving abstraction providers too much control over the language can permit the definition of conflicting  extensions, a problem that is detected ``too late'': when extensions are combined by abstraction clients, who are not necessarily  capable of understanding and fixing the conflict.% Too many constraints can leave it difficult to express real-world abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined. %This requires further research before it would be appropriate to widely rely on them (which we will discuss as we go on).

% \paragraph{Terminology}
% A type system is typically defined in fragments organized around (and often identified by) a \emph{type constructor} (e.g. a fragment defining binary product types, organized around the type constructor $\fvar{prod}$). The syntax of each fragment can be characterized by its contributions to the \emph{concrete syntax} (e.g. expression forms \lstinlinep{(e, e)}, \lstinlinep{e.l} and \lstinlinep{e.r} and type form \lstinlinep{T * T}) and the \emph{abstract syntax}, where terms are written uniformly by applying a \emph{term constructor} to an \emph{index} and zero or more \emph{arguments}, e.g. $\FF{pair}[\FF{triv}](e_1; e_2)$ or $\FF{prj}[\FF{left}](e)$. Term indices are evaluated, if necessary, in an earlier \emph{phase} than term arguments and trivial indices are often omitted for clarity of presentation. The indices of expression terms,  like these examples,  are \emph{static terms}, $\sigma$. \emph{Types}, $\tau$, can be seen as static terms constructed by static term constructor $\FF{ty}$, indexed by a type constructor and having another (possibly trivial) static term as an argument, called (by convention) the \emph{type index}, e.g. $\FF{ty}[\fvar{prod}](\FF{spair}(\tau_1; \tau_2))$, often  abbreviated $\fvar{prod}\texttt{[}\tau_1; \tau_2\texttt{]}$ (%or stylized $\tau_1 \times \tau_2$. %Static terms are also called \emph{type-level terms}.  
%  often, the static terms are not made explicit but rather assumed to be in the ambient metatheory).%Types classify expressions according to the rules comprising the fragment's type system. A type safe language is one where evaluation of a well-typed expression cannot lead to undefined behavior or result in a value of a different type, greatly simplifying static reasoning about program behavior and eliminating the need for many run-time checks.

@@

Our focus in this paper is to enable the introduction of new static type system fragments as safely composable libraries for an existing and widely-used language, leaving its concrete and abstract syntax fixed. This choice is freeing: questions of syntactic ambiguity no longer need to be addressed. Moreover, many tools operate by exhaustive case analysis over the concrete or abstract syntax of the terms of the language they work with (e.g. syntax highlighters, editor modes, pretty-printers, style checkers, documentation generators). These tools will continue to work. %Indeed, enabling the decentralized introduction of new term constructors is the standard example used when describing the more general problem of enabling the extension of a sum type with new cases as well as new  functions over its cases  \cite{Reynolds75}, prompting Wadler to dub it the \emph{expression problem} \cite{wadler1998expression}.

\paragraph{Approach} Our key insight is that few tools operate by exhaustive case analysis over type constructors. We introduce a mechanism that allows library providers to introduce new type system fragments by defining only new type constructors, in libraries, while leaving the concrete and abstract syntax, and the base semantics of the language, fixed.

To support new operations, the semantics of terms are no longer determined directly based on the identity of their term constructors (the usual ``syntax-directed'' approach). Instead, we delegate to a statically defined function associated with a relevant type constructor. The term constructor's identity only serves to determine how the delegate is assigned. We call user-defined type constructors equipped with these static functions and distributed in libraries \emph{active type constructors}.

As a simple example, consider the concrete term \lstinlinep{e.r} (where \lstinlinep{e} is an expression and \lstinlinep{r} is a static label), which maps to the abstract term $\FF{attr}[\text{``r''}](e)$ (i.e. the term constructor $\FF{attr}$, \emph{indexed} by a static representation of the label and taking one \emph{argument}, $e$). A fragment defining statically typed pairs might see this term as referring to projection of the right component of the pair $e$, while a fragment defining functional record types might want it project out a field labeled \verb|r| from $e$, and an object system might introduce a more complex semantics involving dynamic dispatch. The type of dynamically classified Python values would follow Python's own complex field lookup protocol. Were we to naively ``open up'' the syntax and semantics, these fragments would conflict.

Instead, our semantics will delegate responsibility for assigning a type and a translation to this term to a static function associated with the type constructor of the type recursively synthesized by $e$, the \emph{target} of the operation. Thus, tuples, records, objects, dynamic values and variants thereof can  share this syntax.

As we will detail, terms for which there is no clear ``target'' from which to determine a delegate (e.g. number or list literals) instead either use the type they are being analyzed against (following the approach for composable syntax extensions as supported in Wyvern \cite{TSLs}) or, if not available, a ``default'' delegate determined by the type constructor governing the function in which they appear. Symmetric binary terms (e.g. addition) use a third protocol, which we will discuss. There is always exactly one active type constructor delegated  responsibility over a term and this delegate is stable with respect to additional extension imports, so semantic ambiguities cannot arise by construction (unlike systems that treat the type system as a ``bag of rules''). %This is a generalization of a previous mechanism that we have developed, which deals only with extending the elaboration, but not the typechecking, of the bodies of delimited literals \cite{omar-ecoop14}.

>>>>

%\documentclass[12pt]{article}
\documentclass{sigplanconf}
\newcommand{\sigmat}[1]{\sigma_{\text{#1}}}

\newcommand{\F}[1]{{\sf #1}~}
\newcommand{\FF}[1]{{\sf #1}}
\newcommand{\Q}{\FF{Arg}}
\usepackage{sfmath}
\usepackage{times}
%\setlength{\parskip}{0pt}
%\setlength{\parsep}{0pt}
%\setlength{\headsep}{10pt}
%\setlength{\topskip}{0pt}
%\setlength{\topmargin}{0pt}
%\setlength{\topsep}{0pt}
%\setlength{\partopsep}{0pt}

@@

\newcommand{\lip}[1]{\lstinline[language=Python,basicstyle=\ttfamily\small,deletendkeywords={tuple,buffer,map}]{#1}}

@@

%\lefthyphenmin=8
\sloppy
%\makeatletter
%\renewcommand\section{\@startsection{section}{1}{\z@}%
%                       {-18\p@ \@plus -4\p@ \@minus -4\p@}%
%                       {8\p@ \@plus 4\p@ \@minus 4\p@}%
%                       {\normalfont\large\bfseries\boldmath
%                        \rightskip=\z@ \@plus 8em\pretolerance=10000 }}
%\renewcommand\subsection{\@startsection{subsection}{2}{\z@}%
%                       {-18\p@ \@plus -4\p@ \@minus -4\p@}%
%                       {8\p@ \@plus 4\p@ \@minus 4\p@}%
%                       {\normalfont\normalsize\bfseries\boldmath
%                        \rightskip=\z@ \@plus 8em\pretolerance=10000 }}
%\renewcommand\subsubsection{\@startsection{subsubsection}{3}{\z@}%
%                       {-5\p@ \@plus -4\p@ \@minus -4\p@}%
%                       {-0.5em \@plus -0.22em \@minus -0.1em}%
%                       {\normalfont\normalsize\bfseries\boldmath}}
%\renewcommand\paragraph{\@startsection{paragraph}{4}{\z@}%
%                       {-4\p@ \@plus -4\p@ \@minus -4\p@}%
%                       {-0.5em \@plus -0.22em \@minus -0.1em}%
%                       {\normalfont\normalsize\itshape}}
%\makeatother

@@

\conferenceinfo{MODULARITY'15,}{March 16--19, 2015, Fort Collins, CO, USA}
\CopyrightYear{2015}
\crdata{978-1-4503-3249-1/15/03}

\title{{Growing} a Static Type System Inside a Python}
%\subtitle{Unambiguously Growing a Static Type System Inside a Python}
%\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
%\preprintfooter{Distribute within CMU only.}   % 'preprint' option specified.

%\authorinfo{~}{~}{~}
\authorinfo{Cyrus Omar \and Jonathan Aldrich}
          {          Carnegie Mellon University\\
          Pittsburgh, PA, USA}
          {\{comar, aldrich\}@cs.cmu.edu\vspace{-10px}}

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
%gain access to new abstractions, even when they agree that these abstractions
%could be valuable. This suggests a need for languages that satisfy two key
%growth criteria: compatibility with existing programming ecosystems; and
%internal extensibility, so that new abstractions do not require new
%languages.
%We introduce \texttt{typy}, a language compatible with the popular Python ecosystem.
%While Python, like other similar languages, is satisfactory for simple
%scripting, \texttt{typy} is designed for more complex situations where static
%typechecking is beneficial. Unlike most statically-typed languages, \texttt{typy}'s type
%system can be extended from within, by associating compile-time functions with
%type definitions. The compiler selectively invokes these according to a
%type-directed protocol, avoiding interference problems. Despite these design
%constraints, active types in \texttt{typy} can express the static and dynamic semantics
%of a range of functional, object-oriented, parallel and domain-specific
%abstractions, all within libraries.
Programmers prefer using constructs available as libraries for the programming language they already use over those requiring that they switch to a new language dialect. Unfortunately, dialects are seen as the only viable vehicle for constructs that specify non-trivial type systems. %Our aim in this paper is to introduce a mechanism that allows abstraction providers to extend the type system of a language from within, in a way that guarantees that any combination of extensions can be used together without worrying about conflicts. 
Our aim in this paper is to show that there is another way. We 1) introduce a mechanism that gives library providers the ability to introduce new type system fragments modularly; 2) show how this mechanism can itself be implemented as a library, \texttt{typy}, for an existing language, Python; 3) validate our approach with examples of non-trivial type system fragments that are built into other languages but that can be implemented as libraries using this mechanism; 
%, including functional constructs, an object system, a type system for secure string sanitation and a typed foreign function interface to OpenCL
and 4) formally specify the mechanism with a minimal typed lambda calculus, @$\lambda$. The key idea is simple: rather than defining the semantics in the usual syntax-directed manner, we delegate control over the typechecking and translation of each term to some contextually-relevant type constructor. Users  define new type constructors in libraries. %We call type constructors bundled with statically valued functions of this sort \emph{active type constructors}, and this protocol \emph{active typechecking and translation}.%The key idea is to associate semantic logic with type constructors, rather than syntactic forms. 
%Both \texttt{typy} and @$\lambda$ have a fixed syntax. 
%The semantics of each term is determined by delegating control, during the process of typechecking and translating the term to the underlying \emph{internal language}, to methods associated with a relevant \emph{active type constructor}, defined in a library. This process is stable even as additional type constructors are imported, so ambiguities cannot arise. 
% By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction. %In \texttt{typy}, types and functions are defined using Python's syntax, distributed using Python's package system and can call into Python libraries, so \texttt{typy} is uniquely backwards compatible. We show that active type constructors as implemented in \texttt{typy} permit practical library implementations of a variety of type system fragments that would normally need to be built into a language. %We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into \texttt{typy} and then through to OpenCL. % can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library. %We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular. %\texttt{typy} can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces \texttt{typy}, a language that aims to satisfy this more comprehensive set of design criteria. To do so, \texttt{typy} introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of \texttt{typy} is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

\category{D.3.2}{Programming Lan\-guages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
\keywords
type systems; extensible languages; metaprogramming
\section{Introduction}\label{intro}
Convincing programmers to adopt constructs distributed in libraries is easier than con\-vinc\-ing them to use a new programming language. Indeed, re\-cent empirical studies underscore the  difficulty of driving language adoption, finding that extrinsic factors like library compatibility, familiarity and tool support are at least as important to programmers as the intrinsic features of a language \cite{Meyerovich:2013:EAP:2509136.2509515,chen05}.

Despite these steep barriers to adoption, researchers and domain experts continue to design new language dialects, arguing that they have no choice: there often seems to be no other way to faithfully express the constructs that they need, particularly those  specified by a non-trivial static type system. As a concession to concerns about library compatibility, these dialects will often build in a foreign function interface (FFI) to a popular language, like Java. %\emph{Language workbenches} \cite{erdweg2013state} have made it easy to develop sophisticated tooling around such dialects. %As a result, there are now a large number of dialects structured in this manner.

This approach solves the important problem of interfacing with the libraries already written in  this popular language. Interfacing with a library written in a sibling  dialect, however, is difficult. To access such a library via the FFI requires programming against that dialect's implementation, i.e. the code generated by its compiler, rather than its semantics directly. This is anti-modular (if the dialect implementation changes, client code will break), unnatural (syntactic conveniences provided by the dialect are not available) and, for  dialects featuring a type system that statically enforces non-trivial invariants, unsafe. This safety threat can at best be mitigated by including run-time checks in anticipation of this mode of use, though this requires additional effort and comes at the cost of performance.% This likely explains much of the resistance that developers have toward ideas housed in ``toy dialects''. %As more languages are used together, more FFIs need to be constructed and traversed.% This can defeat the productivity gains afforded by these dialects.%To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec. \ref{related-work}).

These problems are severe enough to render even dialects that conservatively introduce only a few new constructs impractical. For example, an empirical study comparing a Java dialect, Habanero-Java (HJ), with a library, \lip{java.util.concurrent}, found that the constructs in HJ were indeed easier to use and provided useful static guarantees \cite{cave2010comparing}. Nevertheless, it concluded that the library-based constructs remained more practical outside the classroom because using HJ, as a distinct dialect of Java with its own type system, would be problematic in settings where some developers had not adopted it. It would also be difficult to use it in combination with other constructs also implemented in dialects of Java. %Moreover, its tool support is more limited.

%These languages become, in essence, merely vehicles for communicating ideas, rather than useful tools.

@@

Language extension mechanisms have promised to mitigate the need for distinct dialects by giving providers the ability to define new syntactic and semantic constructs more granularly, allowing library clients to import them wherever needed. %In   an internally extensible language, programmers  can \emph{granularly} import the primitive constructs best suited to each component of their application or library. Providers can thus more easily develop, deploy and evaluate new abstractions in the context of existing codebases, narrowing one of the gaps between research and practice \cite{basili2008understanding}. 
%Indeed, abstractions could compete on their individual merit, rather than  by the serious adoption barriers mentioned above.
Unfortunately, the mechanisms that are available today have several problems. First, they do not generally preclude the possibility of  syntactic and/or semantic conflicts between separately developed constructs. These conflicts, if detected at all, are detected far too late: when a client happens to use the conflicting constructs within the same program. The client must then patch the problem on their own, once again violating the principle of modularity. The few exceptions to this are mechanisms that support the definition of mutually orthogonal  syntactic sugar, as in Copper \cite{conf/pldi/SchwerdfegerW09} and our recent work on Wyvern \cite{TSLs,sac15}, but these are built atop a fixed semantics. Second, existing mechanisms, including those just mentioned, are often themselves  tied to an obscure dialect or compiler infrastructure and so face the classic ``chicken-and-egg'' problem: they must overcome the same steep barriers to adoption as a dialect like HJ.  %Clients are not generally capable of understanding and fixing the issue productively.% Too many constraints can leave it difficult to express real-world abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined. %This requires further research before it would be appropriate to widely rely on them (which we will discuss as we go on).

% \paragraph{Terminology}
% A type system is typically defined in fragments organized around (and often identified by) a \emph{type constructor} (e.g. a fragment defining binary product types, organized around the type constructor $\fvar{prod}$). The syntax of each fragment can be characterized by its contributions to the \emph{concrete syntax} (e.g. expression forms \lip{(e, e)}, \lip{e.l} and \lip{e.r} and type form \lip{T * T}) and the \emph{abstract syntax}, where terms are written uniformly by applying a \emph{term constructor} to an \emph{index} and zero or more \emph{arguments}, e.g. $\FF{pair}[\FF{triv}](e_1; e_2)$ or $\FF{prj}[\FF{left}](e)$. Term indices are evaluated, if necessary, in an earlier \emph{phase} than term arguments and trivial indices are often omitted for clarity of presentation. The indices of expression terms,  like these examples,  are \emph{static terms}, $\sigma$. \emph{Types}, $\tau$, can be seen as static terms constructed by static term constructor $\FF{ty}$, indexed by a type constructor and having another (possibly trivial) static term as an argument, called (by convention) the \emph{type index}, e.g. $\FF{ty}[\fvar{prod}](\FF{spair}(\tau_1; \tau_2))$, often  abbreviated $\fvar{prod}\texttt{[}\tau_1; \tau_2\texttt{]}$ (%or stylized $\tau_1 \times \tau_2$. %Static terms are also called \emph{type-level terms}.  
%  often, the static terms are not made explicit but rather assumed to be in the ambient metatheory).%Types classify expressions according to the rules comprising the fragment's type system. A type safe language is one where evaluation of a well-typed expression cannot lead to undefined behavior or result in a value of a different type, greatly simplifying static reasoning about program behavior and eliminating the need for many run-time checks.

@@

Our aim in this paper is to take an initial step toward addressing both of these problems by 1) introducing a mechanism that allows library providers to define new type system fragments modularly, and, perhaps most importantly, 2) implementing this mechanism itself as a library, \texttt{typy}, for a widely adopted language, Python. %Programmers can construct types and, using these, define typed functions, from within a Python script. Typed functions, unlike regular Python functions, go through a process of static typechecking and translation. %This process can occur either ahead-of-time or just-in-time.

This design constrains us: we must remain within the confines of Python's existing syntax. But in fact, we find that this constraint is freeing: questions about syntactic ambiguity never arise, the syntax is familiar and tools that require an exhaustive syntax specification (e.g. syntax highlighters, editor plugins, linters and documentation generators)  work without modification. In other words, this design largely sidesteps the \emph{expression problem} \cite{wadler1998expression}.%Taken together, this design addresses many more of the the extrinsic concerns about adoption than a dialect-oriented approach. %Indeed, enabling the decentralized introduction of new term constructors is the standard example used when describing the more general problem of enabling the extension of a sum type with new cases as well as new  functions over its cases  \cite{Reynolds75}, prompting Wadler to dub it the \emph{expression problem} \cite{wadler1998expression}.

We show that you do not need an extensible syntax to support an extensible statically-typed  semantics. The trick is 1) to organize the semantics around a protocol that delegates control over typechecking and translation of each term to logic associated with some contextually-relevant \emph{type constructor} (\emph{tycon}, for brevity); and 2) to give the ability to define new tycons and the logic associated with them to library providers. We call this protocol \textbf{active typechecking and translation} (AT+T).

For example, using \lip{typy}, we will show how a library provider can introduce record types, like those that are built in to languages like ML and Haskell, by defining a tycon called \lip{record}. A different library provider, inspired to give a static semantics to a Javascript-style prototypic object system, can define the active tycon \lip{proto}. We will discuss several other examples throughout the paper.

Types are constructed by applying a tycon to a \emph{type index} using the subscript operator: \lip{tycon[idx]}. For example, we can construct a record type by applying \lip{record} to a mapping from row labels, represented as strings, to types:
\begin{lstlisting}[numbers=none]
R = record[
  'label1' : dyn,    # we discuss dyn and string later
  'label2' : string
]
\end{lstlisting}
Because the language of types, type indices and tycons is Python itself, we are able to use \emph{array slices} to emulate conventional syntax for row specifications, and assignment to define a type synonym, \lip{R}.% We assume that the types \lip{dyn} and \lip{string}, which we will return to shortly, were imported previously.

The following prototypic object type has an analagous field structure (and for simplicity, the default prototype):
\begin{lstlisting}[numbers=none]
P = proto[
  'label1' : dyn,   
  'label2' : string 
]
\end{lstlisting}

The first question that one should ask when shown a new type is ``how do I introduce a value of this type?'' Values of \lip{typy} types are introduced inside \emph{typed functions}, which are those placed under the control of \lip{typy} by the placement of a decorator. Putting the details of typed functions aside for a moment, the form in Python's syntax most similar to the conventional introductory form for records in ML/Haskell is the dictionary literal form, so we would like to be able to introduce values of type \lip{R} inside a typed function using syntax like this:
\begin{lstlisting}[numbers=none]
{'label1': val1, 'label2': val2} 
\end{lstlisting}
But this is also the form that one would hope to use to introduce a value of the prototypic object type \lip{P} above. Moreover, we still need to be able to construct Python dictionaries, of type \lip{dyn}.

Inspired directly by our recent work on \emph{type-specific languages} in Wyvern \cite{TSLs}, we resolve this sort of ambiguity using \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}. When AT+T encounters an expression  of a literal form, it delegates semantic control over it to the tycon of the type that it is being analyzed against. For example, when the expression above appears as an argument to a function \lip{f}, then the argument type of \lip{f} determines its meaning:
\begin{lstlisting}[numbers=none]
f({'label1': val1, 'label2': val2})
\end{lstlisting}
 In positions where there is no expected type, like at the top level of a typed function, the programmer must provide an explicit type ascription. We co-opt Python's array slice syntax for type ascriptions:
\begin{lstlisting}[numbers=none]
myRecord = {'label1': val1, 'label2': val2} [: R]
\end{lstlisting}
Ascriptions can be placed at binding sites as well:
\begin{lstlisting}[numbers=none]
myObj [: P] = {'label1': val1, 'label2': val2}
\end{lstlisting}

The second question one should ask when presented with a new type is ``what can I do with a value of this type?'' For records, the fundamental operation is row projection. Similarly, for prototypic objects, it is field projection. In both cases, the most natural form in Python's syntax for projecting out a row/field labeled \lip{label1} is \lip{e.label1}, where \lip{e} is the record/object, so we must again come up with some way to resolve the ambiguity. Once again, our solution is to take a {tycon-directed} approach: AT+T delegates static control to  the tycon of the type of \lip{e}. %Symmetric binary forms, e.g. addition, have only a slightly more involved delegation protocol.

We will see exactly how  tycons like \lip{record} and \lip{proto} statically control typechecking and translation (to standard Python code) of expressions that they are delegated control over in Sec. \ref{usage}. For now, it suffices to state that the code involved is essentially identical to what one would write if asked to implement a compiler for a dialect that built in the construct in question directly. The novelty of this paper comes not in the constructs themselves, but in how AT+T decentralizes control over the semantics. The locus of control is no longer the language definition but instead individual type constructors, defined in libraries.

% A statically-valued function 
% %, identified based on the name Python's grammar gives to the form, here \lip{syn_Attribute},  %So if \lip{e} is of record type, then the \lip{record} tycon  is delegated control, and if \lip{e} is of prototypic object type then \lip{proto} is delegated control, in either case by calling \lip{syn_Attribute} with a reference to the syntax tree of the term as well as a \emph{typing context}, which we will describe later. 
% must  either synthesize a type for the term as a whole or indicate a type error. If it synthesizes a type, the tycon must then generate a \emph{translation} for the term as a whole, which determines its dynamic behavior. So, for the term \lip{e.label1}, if \lip{e} is of record type, then the \lip{record} tycon is delegated control. It will check that \lip{label1} is one of the row labels specified in the index of the type of \lip{e}. Records can be implemented by translation to tuples (the row labels are only needed statically), so the translation it then generates will project out the appropriate position from the translation recursively generated for \lip{e}.

% We will detail how tycons are implemented as Python classes in Sec. \ref{usage}.  The innovation is in how this protocol decentralizes control over a fixed set of forms, allowing us to define a range of semantic constructs in libraries. The language itself needs to specify only the \emph{delegation protocol} that determines how a tycon is delegated control over each syntactic form.

% More specifically, what this means is that the statically valued function  because that the tycon defines must either synthesize a type for the term as a whole or indicate a type error. If a type is synthesized, the tycon must also generate a \emph{translation} for the term as a whole.

% We call a type constructor equipped with a set of such statically valued functions an \emph{active type constructor}.

% More specifically, the semantics are a \emph{bidirectionally-typed type-directed translation semantics}.
